On the number of tiles visited by a\\ line segment on a rectangular grid

### 2023-04-17 paper.tex
Proofs (hopefully) finished

\newcommand{\mr}{\mathrm}
\newcommand{\st}{\mid}%{\,|\,}
\newcommand{\cond}{\,|\,} % less spacing than \mid
\newcommand{\diff}{\mathrm d}
\let\Re\relax % remove the command...
\DeclareMathOperator{\Re}{Re} % ...to define it differently
\let\Im\relax
\DeclareMathOperator{\Im}{Im}
\DeclareMathOperator{\E}{E}
\newcommand{\funt}{\tau} % function that gives maximum number of touched tiles for real-valued length
\newcommand{\funl}{\lambda} % function that gives infimum real-valued length for a given number of touched tiles
\newcommand{\funti}{T} % function that gives maximum number of touched tiles for integer-valued length
\newcommand{\funli}{\Lambda} % function that gives minimum integer-valued length for a given number of touched tiles
\newcommand{\funta}{\varphi} % function that gives average number of touched tiles for real-valued length
%\newcommand{\funca}{\psi} % function that gives average number of grid line crossings for real-valued length
\newcommand{\probmax}{\rho} % function that gives the probability of achieving the maximum number of touched tiles in the square case for real-valued length
\newcommand{\ras}{\sigma}
\newcommand{\len}{\ell} % length variable, real-valued
\newcommand{\leni}{\ell} % length variable, integer-valued. Same notation, but I keep the different command just in case
\newcommand{\tiles}{t} % number of tiles (integer-valued)
\newcommand{\isolr}{i^+}
\newcommand{\jsolr}{j^+}
\newcommand{\isoli}{i^\ast}
\newcommand{\jsoli}{j^\ast}
\newcommand{\genfun}{q}
\newcommand{\genvar}{s}
\newcommand{\roundterm}{r}
\newcommand{\mss}{M}
\newcommand{\nextrat}{K}
\newcommand{\funcirc}{\Omega}
\newcommand{\funcircesc}{\bar \Omega}

@@

\begin{corollary}
\label{theo: funl, sq, form}
For a unit square grid, and for $\tiles \in \mathbb N$,
\begin{equation}
\funl(\tiles) = \begin{cases}
\displaystyle
0 & \text{for } \tiles =1, 2 \\[1.4mm]
\displaystyle
\frac{\tiles-3}{\sqrt{2}} & \text{for } \tiles \text{ odd, } \tiles \geq 3 \\[4.5mm]
\displaystyle
\frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} {2} & \text{for } \tiles \text{ even, } \tiles \geq 4,
\end{cases}
\end{equation}
or equivalently
\begin{equation}
\label{eq: theo: funt, sq, form; funl}
\funl(\tiles) = \begin{cases}
\displaystyle
0 & \text{for } \tiles =1, 2 \\[1.4mm]
\displaystyle
\sqrt{\left\lceil \frac{(\tiles-3)^2} {2} \right\rceil} & \text{for } \tiles \geq 3.
\end{cases}
\end{equation}
\end{corollary}
% Theorem~\ref{theo: funl, form} for $a=b=1$ gives $\sqrt{\frac{(\tiles-3)^2+1} {2}}$. Instead, this corollary uses the equivalent expression $\frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} {2}$, which is better for subsequent analysis.

@@

\begin{theorem}
\label{theo: funli}
For $\tiles \in \mathbb N$,
% $\funli(\tiles)$ can be computed as
\begin{equation}
\label{eq: funli}
\funli(\tiles) = \begin{cases}
\displaystyle
1 & \text{for } \tiles \leq 3 \\[1.3mm]
\displaystyle
\left \lceil \sqrt{\frac{(\tiles-3)^2} 2 + 1} \ \right \rceil & \text{for } \tiles \geq 4.
\end{cases}
\end{equation}
In addition,
\begin{enumerate}
\item
This sequence is non-decreasing. Except for the initial run of $3$ equal values, it is formed by runs of $1$ or $2$ equal values, with an increment of $1$ between consecutive runs.
\item
There can be no more than $3$ consecutive terms that are different.
\item
A run of $2$ equal values always has $2$ different terms before and $2$ different terms after the run, except for the initial terms $1, 1, 1, 2, 2, 3, 3$.
\end{enumerate}
\end{theorem}

@@

\begin{proposition}
\label{prop: Pr i}
Given $a, b, \len \in \mathbb R^+$, consider a grid with parameters $a, b$ and a uniformly random segment of length $\len$, as defined above. Let the random variables $i$, $j$ represent the normalized dimensions of the discrete bounding rectangle. For $n \in \mathbb N$,
\begin{equation}
\label{eq: Pr i geq n}
\Pr[i \geq n] = \begin{cases}
\displaystyle
1 &\text{if\ \ } \displaystyle n =1 \\[1 mm]
\displaystyle
\frac{2\len}{\pi a} \left(f\left(\frac{a(n-1)}{\len}\right)-f\left(\frac{a(n-2)}{\len} \right)\right) &\text{if\ \ } \displaystyle 2 \leq n < \frac \len a + 1 \\[4 mm]
\displaystyle
\frac{2\len}{\pi a}\,\, \left(1 - f\left(\frac{a(n-2)}{\len}\right)\right) &\text{if\ \ } \displaystyle \frac\len a + 1 \leq n < \frac\len a+2 \\[3 mm]
\displaystyle
0 &\text{if\ \ } \displaystyle \frac \len a + 2 \leq  n; \\
\end{cases}
\end{equation}
and $\Pr[j \geq n]$ is given by the same expressions with $a$ replaced by $b$.
\end{proposition}

@@

The probability of visiting the maximum number of tiles for a rectangular grid is difficult to compute, due to the irregularity of the optimal $(i,j)$ pairs (which was discussed in \S\ref{part: fund results}; in particular see Figure~\ref{fig: pairs_bounds}). In a square grid, however, the simplicity of the set of optimal pairs makes the problem more tractable.

@@

\begin{theorem}
\label{theo: probmax, sq}
Given $\len \in \mathbb R^+$, consider a unit square grid and a uniformly random segment of length $\len$, as previously defined. The probability $\probmax(\len)$ that the segment visits the maximum number of tiles is
\begin{equation}
\label{eq: probmax impl}
\probmax(\len) = \sum_{\substack{i,j \geq 2,\ % 
i+j-1=\tiles,\\
(i-2)^2+(j-2)^2<\len^2
}} g\left(\len, i-2, j-2 \right),
\end{equation}
or equivalently, defining $\nextrat = \sqrt{\len^2/2-(\tiles-3)^2/4}$,
\begin{equation}
\label{eq: probmax expl}
\probmax(\len) =
\begin{cases}
\displaystyle
g\left(\len, \frac{\tiles-3} 2, \frac{\tiles-3} 2 \right) +
2 \sum_{k=1}^{\lceil\nextrat-1 \rceil} g\left(\len, \frac{\tiles-3} 2 - k, \frac{\tiles-3} 2 + k \right) &\text{for $\tiles$ odd} \\[6mm] % *!*
\displaystyle
2 g\left(\len, \frac{\tiles-4} 2, \frac{\tiles-2} 2 \right) +
2 \sum_{k=1}^{\lceil\nextrat-3/2 \rceil} g\left(\len, \frac{\tiles-4} 2 - k, \frac{\tiles-2} 2 + k \right) &\text{for $\tiles$ even}.
\end{cases}
\end{equation}
\end{theorem}

@@

Consider $\len \in \mathbb R^+$, and let $\tiles = \funt(\len)$. By Proposition~\ref{prop: i+j-1}, the segment visits $\tiles$ tiles almost surely if the dimensions $i$, $j$ of its discrete bounding rectangle satisfy $i+j-1=\tiles$. In addition, a segment that visits the maximum number of tiles always has $i, j \geq 2$, because if $i=1$ the segment can always be shifted and possibly tilted to cross a vertical grid line, thus increasing $i$ to $2$, without reducing $j$. The reasoning for $j=1$ is analogous. Thus for $i, j \geq 2$, $i+j-1=\tiles$ let $\probmax_{i,j}(\len)$ denote the probability that the segment has a discrete bounding rectangle with dimensions $i, j$. Then
\begin{equation}
\label{eq: probmax len probmax i,j len}
\probmax(\len) =  \sum_{i,j \geq 2,\ i+j-1=\tiles} \probmax_{i,j}(\len).
\end{equation}

For any $i, j \geq 2$ with $i+j-1=\tiles$, the segment has a discrete bounding rectangle with dimensions $i, j$ if and only if its second endpoint is in the tile with lower-left corner $(i-1,j-1)$ or in one of the other three symmetrical tiles with respect to the lines $x=1/2$ and $y=1/2$, i.e.~those with lower-left corners $(i-1,1-j)$, $(1-i,1-j)$ and $(1-i,1-j)$. An example is shown in Figure~\ref{fig: probmax_symmetric_tiles}. Consequently, $\probmax_{i,j}(\len)$ can be obtained by computing the probability that the second endpoint is in the tile with lower-left corner $(i-1,j-1)$ and then multiplying by $4$.

\begin{figure}
\centering%
\includegraphics[width=.5\textwidth]{probmax_symmetric_tiles_}%
\caption{Symmetric tiles for the computation of $\probmax_{i,j}(\len)$ on a unit square grid. Example for $i=3$, $j=4$%
}%
\label{fig: probmax_symmetric_tiles}%
\end{figure}%

@@

For a fixed orientation $\theta$, the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ if and only if shifting the segment up and to the right so that its first endpoint is at $(1,1)$ results in the second endpoint still being in the tile with lower-left corner $(i-1,j-1)$, as shown in Figure~\ref{fig: probmax_funt_odd}. It cannot be the case that the segment ``overshoots'' past this tile, because that would yield a number of visited tiles exceeding $\tiles$, which is the maximum. In other words, only one corner of the tile can be contained within the arc in the figure, namely its lower-left corner $(i-1,j-1)$.

\begin{figure}
\centering
\includegraphics[width=.56\textwidth]{probmax_funt_odd}%
\caption{Reachability from the reference tile for a given $\tiles = \funt(\len)$ on a unit square grid. Example for $\tiles = 7$
}%
\label{fig: probmax_funt_odd}%
\end{figure}%

There is a range of values of $\theta$ for which the above condition is satisfied. This range corresponds to the part of the arc with solid line in Figure~\ref{fig: probmax_funt_odd}. For each $\theta$ in the allowed range, the valid positions for the second endpoint of the segment are in a rectangle contained in the tile with lower-left corner $(i-1,j-1)$. These are the positions that cause the first endpoint to be in the reference tile. The rectangle associated with an example $\theta$ is shown in Figure~\ref{fig: probmax_funt_odd}, shaded. The area of this rectangle is the probability $\probmax_{i, j}(\len,\theta)$ that the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ conditioned on $\theta$. As seen in the figure, this area is
\begin{equation}
\label{eq: probmax: probmax theta odd}
\probmax_{i, j}(\len, \theta) = \left(\len\cos\theta - i+2\right) \left(\len\sin\theta - j + 2\right).
\end{equation}
The range of allowed values for $\theta$, as can also be deduced from the figure, is $(\theta_0, \theta_1)$ with
\begin{align}
\label{eq: odd theta 0}
\theta_0 & = \arcsin\frac{j-2}{\len}, \\
\label{eq: odd theta 1}
\theta_1 & = \arccos\frac{i-2}{\len}.
\end{align}
For $\theta$ outside of this range the conditional probability $\probmax_{i, j}(\len, \theta)$ is $0$. Averaging $\probmax_{i, j}(\len, \theta)$ over $\theta$ gives the probability that the segment visits this tile, and then multiplying by $4$ yields $\probmax_{i.j}(\len)$:
\begin{equation}
\begin{split}
\label{eq: probmax: cond on theta}
\probmax_{i,j}(\len) &= 4 \int_{\theta_0}^{\theta_1} \frac{\probmax_{i, j}(\len, \theta)}{2\pi} \, \diff \theta \\
&= \frac 2 {\pi} \int_{\arcsin\frac{j-2}{\len}}^{\arccos\frac{i-2}{\len}} \left( \len\cos\theta - i + 2 \right)  \left( \len\sin\theta - j + 2 \right) \, \diff \theta
= g( \len, i-2, j-2).
\end{split}
\end{equation}

The above arguments show that $\probmax_{i,j}(\len)$ is given by \eqref{eq: probmax: cond on theta} if the condition \eqref{eq: i-2 j-2 len} holds, and equals $0$ otherwise. Using this into \eqref{eq: probmax len probmax i,j len} yields \eqref{eq: probmax impl} .

Given $\tiles$, the range of lengths $\len$ such that $\tiles = \funt(\len)$ is either \eqref{eq: square, odd max tiles, lengths} or \eqref{eq: square, even max tiles, lengths}. For $\len$ in this range, the number of tiles with lower-left corners $(i-1,j-1)$, $i,j \geq 2$, $i+j-1=\tiles$ that can be reached, i.e.~the number of terms in \eqref{eq: probmax impl}, depends on $\len$. If $\tiles$ is odd the minimum number is $1$, corresponding to a tile with lower-left corner on the line $x=y$, which gives the first term in the first expression in \eqref{eq: probmax expl}. For $\tiles$ even the minimum number is $2$, corresponding to two symmetrical tiles whose $x$ and $y$ coordinates differ by $1$, which give the first term in the second expression. Either for $\tiles$ odd or even, additional tiles may become reachable as $\len$ grows (within the range allowed by $\tiles$), and these always occur in symmetrical pairs. These tiles (if any) correspond to the terms in the (possibly empty) sum indexed by $k$ in either expression in \eqref{eq: probmax expl}.

@@

As can be seen in Figure~\ref{fig: probmax_terms}, for each $\tiles$ there is a maximum number of tiles that can be reached, beyond which incrementing $\len$ only results in $\tiles$ increasing by $1$. For a given $\len$, the number of reachable tiles can be determined by counting how many markers are covered by the length $w$ defined in Figure~\ref{fig: probmax_nterms}. This length satisfies the relationship $w^2 = \len^2 - (\tiles-3)^2/2$, and thus $K$ as defined in the theorem equals $w/\sqrt{2}$. From the figure, the number of terms in the sum over $k$ (empty circular markers) is $\lceil w/\sqrt 2 - 1 \rceil$ or $\lceil w/\sqrt 2 - 3/2 \rceil$ for $\tiles$ odd or even, i.e.~$\lceil \nextrat - 1 \rceil$ or $\lceil \nextrat - 3/2 \rceil$ respectively. This establishes \eqref{eq: probmax expl}.
\begin{figure}
\centering%
\subfigure[Odd $\tiles$]{%
\label{fig: probmax_nterms_odd}%
\includegraphics[width=.65\textwidth]{probmax_nterms_odd}%
}\\%
\subfigure[Even $\tiles$]{%
\label{fig: probmax_nterms_even}%
\includegraphics[width=.65\textwidth]{probmax_nterms_even}%
}%
\caption{Computation of the number of reachable tiles with lower-left corners $(i-1,j-1)$, $i+j-1=\tiles=\funt(\len)$ on a unit square grid}%
\label{fig: probmax_nterms}%
\end{figure}%
\end{proof}

The technique used in the proof Theorem~\ref{theo: probmax, sq} could in principle be used for obtaining the probability that the number of tiles visited by the segment equals or exceeds any given value $\tiles < \funt(\len)$. However, the process is more cumbersome. Specifically, with reference to Figure~\ref{fig: probmax_funt_odd}, the arc in this more general setting can enclose any number of corners of the considered tile, not necessarily $1$, which makes the characterization of the shaded rectangles more complicated.
% because several cases need to be distinguished

\begin{figure}%
\centering%
\includegraphics[width=.75\textwidth]{probmax_len}%
\caption{Probability $\probmax(\len)$ that a random segment of length $\len$ visits the maximum number of tiles on a unit square grid%
}%
\label{fig: probmax_len}%
\end{figure}%

The probability $\probmax(\len)$ of visiting the maximum number of tiles on a unit square grid, computed from Theorem~\ref{theo: probmax, sq}, is shown in Figure~\ref{fig: probmax_len} shows. As $\len$ grows, $\probmax(\len)$ has a jump when $\funt(\len)$ increases by $1$ (and there $\probmax(\len)$ takes the value approached from the left). This happens when $\len$ equals the right endpoint of the interval \eqref{eq: square, odd max tiles, lengths} for $\tiles$ odd or of the interval \eqref{eq: square, even max tiles, lengths} for $\tiles$ even. These length values will be denoted as $\len_\tiles$:
\begin{equation}
\label{eq: len tiles}
\len_\tiles = \begin{cases}
\displaystyle
\frac{\sqrt{(\tiles-3)^2 + (\tiles-1)^2}} {2} = \sqrt{\frac{(\tiles-2)^2+1}{2}} & \text{ for $\tiles$ odd, $\tiles \geq 3$} \\
\displaystyle
\frac{\tiles-2}{\sqrt{2}} & \text{ for $\tiles$ even, $\tiles \geq 4$},
\end{cases}
\end{equation}
and satisfy $\funt(\len_\tiles) = \tiles$.

The first continuous section plotted in Figure~\ref{fig: probmax_len} corresponds to a maximum number of visited tiles $\tiles=3$, for lengths in the interval $(0, \len_3]$, where $\len_3= 1$. The second corresponds to $\tiles=4$, for lengths in $(\len_3, \len_4]$, where $\len_4= \sqrt{2}$. Within each continuous section the probability monotonically increases from $0$ to a maximum value. The heights of the maxima are asymptotically proportional to $1/\sqrt{\len_\tiles}$, as established by the next result.

>>>>

\newcommand{\mr}{\mathrm}
\newcommand{\st}{\mid}%{\,|\,}
\newcommand{\cond}{\,|\,} % less spacing than \mid
\newcommand{\diff}{\mathrm d}
\let\Re\relax % remove the command...
\DeclareMathOperator{\Re}{Re} % ...to define it differently
\let\Im\relax
\DeclareMathOperator{\Im}{Im}
\DeclareMathOperator{\E}{E}
\newcommand{\funt}{\tau} % function that gives maximum number of touched tiles for real-valued length
\newcommand{\funl}{\lambda} % function that gives infimum real-valued length for a given number of touched tiles
\newcommand{\funti}{T} % function that gives maximum number of touched tiles for integer-valued length
\newcommand{\funli}{\Lambda} % function that gives minimum integer-valued length for a given number of touched tiles
\newcommand{\funta}{\varphi} % function that gives average number of touched tiles for real-valued length
%\newcommand{\funca}{\psi} % function that gives average number of grid line crossings for real-valued length
\newcommand{\probmax}{\rho} % function that gives the probability of achieving the maximum number of touched tiles in the square case for real-valued length
\newcommand{\ras}{\sigma}
\newcommand{\len}{\ell} % length variable, real-valued
\newcommand{\leni}{\ell} % length variable, integer-valued. Same notation, but I keep the different command just in case
\newcommand{\tiles}{t} % number of tiles (integer-valued)
\newcommand{\isolr}{i^+}
\newcommand{\jsolr}{j^+}
\newcommand{\isoli}{i^\ast}
\newcommand{\jsoli}{j^\ast}
\newcommand{\genfun}{q}
\newcommand{\genvar}{s}
\newcommand{\roundterm}{r}
\newcommand{\mss}{M}
%\newcommand{\nextrat}{K}
\newcommand{\funcirc}{\Omega}
\newcommand{\funcircesc}{\bar \Omega}
\newcommand{\fdp}{p}

@@

\begin{corollary}
\label{theo: funl, sq, form}
For a unit square grid, and for $\tiles \in \mathbb N$,
\begin{equation}
\funl(\tiles) = \begin{cases}
\displaystyle
0 & \text{for } \tiles =1, 2 \\[1.4mm] % *!*
\displaystyle
\frac{\tiles-3}{\sqrt{2}} & \text{for } \tiles \text{ odd, } \tiles \geq 3 \\[4.5mm] % *!*
\displaystyle
\frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} {2} & \text{for } \tiles \text{ even, } \tiles \geq 4,
\end{cases}
\end{equation}
or equivalently
\begin{equation}
\label{eq: theo: funt, sq, form; funl}
\funl(\tiles) = \begin{cases}
\displaystyle
0 & \text{for } \tiles =1, 2 \\[1.4mm]  % *!*
\displaystyle
\sqrt{\left\lceil \frac{(\tiles-3)^2} {2} \right\rceil} & \text{for } \tiles \geq 3.
\end{cases}
\end{equation}
\end{corollary}
% Theorem~\ref{theo: funl, form} for $a=b=1$ gives $\sqrt{\frac{(\tiles-3)^2+1} {2}}$. Instead, this corollary uses the equivalent expression $\frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} {2}$, which is better for subsequent analysis.

@@

\begin{theorem}
\label{theo: funli}
For $\tiles \in \mathbb N$,
% $\funli(\tiles)$ can be computed as
\begin{equation}
\label{eq: funli}
\funli(\tiles) = \begin{cases}
\displaystyle
1 & \text{for } \tiles \leq 3 \\[1.3mm] % *!*
\displaystyle
\left \lceil \sqrt{\frac{(\tiles-3)^2} 2 + 1} \ \right \rceil & \text{for } \tiles \geq 4.
\end{cases}
\end{equation}
In addition,
\begin{enumerate}
\item
This sequence is non-decreasing. Except for the initial run of $3$ equal values, it is formed by runs of $1$ or $2$ equal values, with an increment of $1$ between consecutive runs.
\item
There can be no more than $3$ consecutive terms that are different.
\item
A run of $2$ equal values always has $2$ different terms before and $2$ different terms after the run, except for the initial terms $1, 1, 1, 2, 2, 3, 3$.
\end{enumerate}
\end{theorem}

@@

\begin{proposition}
\label{prop: Pr i}
Given $a, b, \len \in \mathbb R^+$, consider a grid with parameters $a, b$ and a uniformly random segment of length $\len$, as defined above. Let the random variables $i$, $j$ represent the normalized dimensions of the discrete bounding rectangle. For $n \in \mathbb N$,
\begin{equation}
\label{eq: Pr i geq n}
\Pr[i \geq n] = \begin{cases}
\displaystyle
1 &\text{if\ \ } \displaystyle n =1 \\[1 mm] % *!*
\displaystyle
\frac{2\len}{\pi a} \left(f\left(\frac{a(n-1)}{\len}\right)-f\left(\frac{a(n-2)}{\len} \right)\right) &\text{if\ \ } \displaystyle 2 \leq n < \frac \len a + 1 \\[4 mm] % *!*
\displaystyle
\frac{2\len}{\pi a}\,\, \left(1 - f\left(\frac{a(n-2)}{\len}\right)\right) &\text{if\ \ } \displaystyle \frac\len a + 1 \leq n < \frac\len a+2 \\[3 mm] % *!*
\displaystyle
0 &\text{if\ \ } \displaystyle \frac \len a + 2 \leq  n;
\end{cases}
\end{equation}
and $\Pr[j \geq n]$ is given by the same expressions with $a$ replaced by $b$.
\end{proposition}

@@

The probability of visiting the maximum number of tiles for a rectangular grid is difficult to compute. The reason is analogous to the irregularity of the optimal $(i,j)$ pairs which was discussed in \S\ref{part: fund results}. In a square grid, however, the problem is more tractable.

@@

\begin{theorem}
\label{theo: probmax, sq}
Given $\len \in \mathbb R^+$, consider a unit square grid and a uniformly random segment of length $\len$, as previously defined. The probability $\probmax(\len)$ that the segment visits the maximum number of tiles $\tiles= \funt(\len)$ is
\begin{equation}
\label{eq: probmax impl}
\probmax(\len) = \sum_{\substack{i,j \geq 2,\ % 
i+j-1=\tiles,\\
(i-2)^2+(j-2)^2<\len^2
}} g\left(\len, i-2, j-2 \right),
\end{equation}
or equivalently, defining $w = \sqrt{\len^2-(\tiles-3)^2/2}$,
\begin{equation}
\label{eq: probmax expl}
\probmax(\len) =
\begin{cases}
\displaystyle
g\left(\len, \frac{\tiles-3} 2, \frac{\tiles-3} 2 \right) +
2 \sum_{k=1}^{\left\lceil \frac w {\sqrt{2}} - 1 \right\rceil} g\left(\len, \frac{\tiles-3} 2 - k, \frac{\tiles-3} 2 + k \right) &\text{for $\tiles$ odd} \\[6mm] % *!*
\displaystyle
2 g\left(\len, \frac{\tiles-4} 2, \frac{\tiles-2} 2 \right) +
2 \sum_{k=1}^{\left\lceil \frac w {\sqrt{2}} - \frac 3 2 \right\rceil} g\left(\len, \frac{\tiles-4} 2 - k, \frac{\tiles-2} 2 + k \right) &\text{for $\tiles$ even}.
\end{cases}
\end{equation}
\end{theorem}

@@

Consider $\len \in \mathbb R^+$, and let $\tiles = \funt(\len)$. By Proposition~\ref{prop: i+j-1}, the segment visits $\tiles$ tiles almost surely if the dimensions $i$, $j$ of its discrete bounding rectangle are such that $i+j-1=\tiles$. In addition, a segment that visits the maximum number of tiles always has $i, j \geq 2$. Namely, if $i=1$ the segment can always be shifted and possibly tilted to cross a vertical grid line, thus increasing $i$ to $2$, without reducing $j$. The reasoning for $j=1$ is analogous. Thus for $i, j \geq 2$, $i+j-1=\tiles$ let $\probmax_{i,j}(\len)$ denote the probability that the segment has a discrete bounding rectangle with dimensions $i, j$. Then
\begin{equation}
\label{eq: probmax len probmax i,j len}
\probmax(\len) =  \sum_{i,j \geq 2,\ i+j-1=\tiles} \probmax_{i,j}(\len).
\end{equation}

For any $i, j \geq 2$ with $i+j-1=\tiles$, the segment has a discrete bounding rectangle with dimensions $i, j$ if and only if its second endpoint is in the tile with lower-left corner $(i-1,j-1)$ or in one of the other three symmetrical tiles with respect to the lines $x=1/2$ and $y=1/2$, i.e.~those with lower-left corners $(i-1,1-j)$, $(1-i,1-j)$ and $(1-i,1-j)$. An example is shown in Figure~\ref{fig: probmax_symmetric_tiles}. Consequently, $\probmax_{i,j}(\len)$ can be obtained by computing the probability that the second endpoint is in the tile with lower-left corner $(i-1,j-1)$ and multiplying by $4$.

\begin{figure}
\centering%
\includegraphics[width=.53\textwidth]{probmax_symmetric_tiles}%
\caption{Symmetrical tiles for the computation of $\probmax_{i,j}(\len)$ on a unit square grid. Example for $i=3$, $j=4$%
}%
\label{fig: probmax_symmetric_tiles}%
\end{figure}%

@@

For a fixed orientation $\theta$, the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ if and only if shifting the segment up and to the right so that its first endpoint is at $(1,1)$ results in the second endpoint still being in the tile with lower-left corner $(i-1,j-1)$. This is illustrated in Figure~\ref{fig: probmax_funt_odd}. It cannot be the case that the segment ``overshoots'' past this tile, because that would yield a number of visited tiles exceeding $\tiles$, which is the maximum. In other words, only one corner of the tile can be contained within the arc in the figure, namely its lower-left corner $(i-1,j-1)$.

\begin{figure}
\centering
\includegraphics[width=.56\textwidth]{probmax_funt_odd}%
\caption{Segment orientations and positions for a given $\tiles = \funt(\len)$ on a unit square grid. Example for $\tiles = 7$
}%
\label{fig: probmax_funt_odd}%
\end{figure}%

There is a range of values of $\theta$ for which the above condition is satisfied. This corresponds to the part of the arc with solid line in Figure~\ref{fig: probmax_funt_odd}. For each $\theta$ in this range, the valid positions for the second endpoint of the segment are in a rectangle contained in the tile with lower-left corner $(i-1,j-1)$. These are the positions that cause the first endpoint to be in the reference tile. The rectangle associated with an example $\theta$ is shown in Figure~\ref{fig: probmax_funt_odd}, shaded. The area of this rectangle is the probability $\probmax_{i, j}(\len,\theta)$ that the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ conditioned on $\theta$. As seen in the figure, this area is
\begin{equation}
\label{eq: probmax: probmax theta odd}
\probmax_{i, j}(\len, \theta) = \left(\len\cos\theta - i+2\right) \left(\len\sin\theta - j + 2\right).
\end{equation}
The range of allowed values for $\theta$, as can also be deduced from the figure, is $(\theta_0, \theta_1)$ with
\begin{align}
\label{eq: odd theta 0}
\theta_0 & = \arcsin\frac{j-2}{\len}, \\
\label{eq: odd theta 1}
\theta_1 & = \arccos\frac{i-2}{\len}.
\end{align}
For $\theta$ outside of this range the conditional probability $\probmax_{i, j}(\len, \theta)$ is $0$. Averaging $\probmax_{i, j}(\len, \theta)$ over $\theta$ gives the probability that the segment visits this tile, and then multiplying by $4$ yields $\probmax_{i.j}(\len)$:
\begin{equation}
\begin{split}
\label{eq: probmax: cond on theta}
\probmax_{i,j}(\len) &= 4 \int_{\theta_0}^{\theta_1} \frac{\probmax_{i, j}(\len, \theta)}{2\pi} \, \diff \theta \\
&= \frac 2 {\pi} \int_{\arcsin\frac{j-2}{\len}}^{\arccos\frac{i-2}{\len}} \left( \len\cos\theta - i + 2 \right)  \left( \len\sin\theta - j + 2 \right) \, \diff \theta
= g( \len, i-2, j-2).
\end{split}
\end{equation}

The above arguments show that $\probmax_{i,j}(\len)$ is given by \eqref{eq: probmax: cond on theta} if the condition \eqref{eq: i-2 j-2 len} holds, and equals $0$ otherwise. Using this into \eqref{eq: probmax len probmax i,j len} yields \eqref{eq: probmax impl}.

Given $\tiles$, the range of lengths $\len$ such that $\tiles = \funt(\len)$ is either \eqref{eq: square, odd max tiles, lengths} or \eqref{eq: square, even max tiles, lengths}. For $\len$ in this range, the number of reachable tiles with lower-left corners $(i-1,j-1)$, $i,j \geq 2$, $i+j-1=\tiles$, i.e.~the number of terms in \eqref{eq: probmax impl}, depends on $\len$. If $\tiles$ is odd the minimum number is $1$, corresponding to a tile with its lower-left corner on the line $x=y$, which gives the first term in the first expression in \eqref{eq: probmax expl}. For $\tiles$ even the minimum number is $2$, corresponding to two symmetrical tiles whose lower-left $x$ and $y$ coordinates differ by $1$, which give the first term in the second expression. Either for $\tiles$ odd or even, additional tiles may become reachable as $\len$ grows (within the range allowed by $\tiles$), and these always occur in symmetrical pairs. These tiles (if any) correspond to the terms in the (possibly empty) sum indexed by $k$ in either expression in \eqref{eq: probmax expl}.

@@

As can be seen in Figure~\ref{fig: probmax_terms}, for each $\tiles$ there is a maximum number of tiles that can be reached, beyond which incrementing $\len$ only results in $\tiles$ increasing by $1$. For a given $\len$, the number of reachable tiles can be determined by counting how many markers are covered by the stretch $w$ defined in Figure~\ref{fig: probmax_nterms}. This satisfies the relationship $w^2 = \len^2 - (\tiles-3)^2/2$. From the figure, the number of terms in the sum over $k$ (empty circular markers) is $\lceil w/\sqrt 2 - 1 \rceil$ for $\tiles$ odd and $\lceil w/\sqrt 2 - 3/2 \rceil$ for $\tiles$ even. This establishes \eqref{eq: probmax expl}.

\begin{figure}
\centering%
\subfigure[Odd $\tiles$]{%
\label{fig: probmax_nterms_odd}%
\includegraphics[width=.65\textwidth]{probmax_nterms_odd}%
}\\%
\subfigure[Even $\tiles$]{%
\label{fig: probmax_nterms_even}%
\includegraphics[width=.65\textwidth]{probmax_nterms_even}%
}%
\caption{Computation of the number of reachable tiles with lower-left corners $(i-1,j-1)$, $i+j-1=\tiles=\funt(\len)$ on a unit square grid}%
\label{fig: probmax_nterms}%
\end{figure}%

It should be noted that, in principle, the technique used in this proof could be employed for obtaining the probability that the number of tiles visited by the segment equals or exceeds any given value $\tiles < \funt(\len)$. However, the process is more cumbersome. Specifically, with reference to Figure~\ref{fig: probmax_funt_odd}, the arc in this more general setting can enclose any number of corners of the considered tile, not necessarily $1$, which makes the characterization of the shaded rectangles more complicated.
\end{proof}

The probability $\probmax(\len)$ of visiting the maximum number of tiles on a unit square grid, computed from Theorem~\ref{theo: probmax, sq}, is shown in Figure~\ref{fig: probmax_len}. As $\len$ grows, $\probmax(\len)$ has a jump when $\funt(\len)$ increases by $1$. This happens when $\len$ equals the right endpoint of the interval \eqref{eq: square, odd max tiles, lengths} for $\tiles$ odd or of the interval \eqref{eq: square, even max tiles, lengths} for $\tiles$ even. These length values will be denoted as $\len_\tiles$:
\begin{equation}
\label{eq: len tiles}
\len_\tiles = \begin{cases}
\displaystyle
\frac{\sqrt{(\tiles-3)^2 + (\tiles-1)^2}} {2} = \sqrt{\frac{(\tiles-2)^2+1}{2}} & \text{ for $\tiles$ odd, $\tiles \geq 3$} \\[2 mm] % *!*
\displaystyle
\frac{\tiles-2}{\sqrt{2}} & \text{ for $\tiles$ even, $\tiles \geq 4$}.
\end{cases}
\end{equation}
Clearly, $\funt(\len_\tiles) = \tiles$ and $\lim_{\len \rightarrow \len_\tiles^- } \probmax(\len) = \probmax(\len_\tiles)$.

\begin{figure}%
\centering%
\includegraphics[width=.74\textwidth]{probmax_len}%
\caption{Probability $\probmax(\len)$ that a random segment of length $\len$ visits the maximum number of tiles on a unit square grid%
}%
\label{fig: probmax_len}%
\end{figure}%

The first continuous section seen in Figure~\ref{fig: probmax_len} corresponds to a maximum number of visited tiles $\tiles=3$, for lengths in the interval $(0, \len_3]$, where $\len_3= 1$. The second corresponds to $\tiles=4$, for lengths in $(\len_3, \len_4]$, where $\len_4= \sqrt{2}$. Within each continuous section the probability monotonically increases from $0$ to a maximum value. The heights of the maxima are asymptotically proportional to $1/\sqrt{\len_\tiles}$, as established by the next result.
